Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-17 paper.tex
Reorganized the introduction to be a bit less technical and a bit better (I hope) organized from the broad to the specific.  Added some more classic references to go with the broader context.  Let me know what you think...

\title{Cooling Requirements for the Vertical Shear Instability in
  astrophysical disks \\ and Implications for Protoplanetary Disks}
%\title{Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula} 
\author{Min-Kai Lin \& Andrew Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

\begin{abstract}
  It is difficult to understand how cold circumstellar disks accrete onto their central stars.  Due to 
  lower levels of ionization, the magnetic transport mechanisms that operate in warmer disks are less effective. 
  A hydrodynamic mechanism, the vertical shear instability (VSI), offers a means to drive angular momentum 
  transport in cold accretion disks such as protoplanetary disks.  
  The VSI is driven by a weak vertical gradient in the disk's orbital motion.
  In order to grow, the VSI must overcome vertical buoyancy, a strong stabilizing influence in cold disks, where 
  heating is dominated by external irradiation.  Rapid cooling, via radiative losses, reduces the effective buoyancy
  and allows the VSI to operate.  In this paper, we quantify the cooling timescale, $t_c$, needed for growth of the VSI
  and consider the implications for protoplanetary disks where $t_c$ is determined by the opacity of dust 
  grains.  We perform a linear analysis of the VSI with cooling in vertically global and radially local disk models.
  For irradiated disks, we find that the VSI is most vigorous for rapid cooling with $t_c < \Omega_K^{-1} \epsilon |q| / (\gamma -1)$
  in terms of the Keplerian orbital frequency, $\OmK$; the disk's aspect ratio, $\epsilon \ll 1$; the radial 
  powerlaw temperature gradient, $q$; and the adiabatic index, $\gamma$.  For longer cooling times, the VSI 
  is much less effective because growth slows and shifts to smaller length scales, which are more prone to 
  viscous or turbulent decay.  We apply our results to a  Minimum Mass Solar Nebula (MMSN) disk model.  We
  find that the VSI is most effective at intermediate radii, from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.  Growth is suppressed by long cooling times both in the opaque 
  inner disk and the optically thin outer disk.  A reduction in the dust opacity by a factor of 10 increases cooling times 
  enough to quench the VSI at all disk radii.  Thus the formation of solid protoplanets, a sink for dust grains, can impede the VSI.

%  A robust hydrodynamic mechanism for turbulent angular momentum
%  transport is important for cold accretion disks such as 
%  protoplanetary disks. The vertical shear instability (VSI) has 
%  recently been suggested to provide such a pathway, since it operates in
%  disks with rotation profiles that depend on the height from the 
%  disk midplane, which is generally expected in  
%  astrophysical disks. However, vertical shear is weak in thin disks, 
%  so the VSI can easily be stabilized by buoyancy forces in the
%  presence of a stable vertical entropy  gradient. % , which is also
%  % expected in astrophysical disks. 
%  Consequently,  the VSI
%  requires a rather short thermal or cooling timescale in
%  sub-adabatically stratified disks.      
%  A quantitative examination of this thermodynamic 
%  requirement is needed to assess the viability of
%  the VSI in realistic disks. To this end,  
%  we study the linear, axisymmetric VSI in radially local, 
%  vertically global models of stably stratified accretion disks in which the disk  
%  temperature $T$ is relaxed to its initial value over a characteristic timescale 
%  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
%  and $\Omega_k$ is the Keplerian rotation frequency. % We explore how
%  % the VSI depends on the thermodynamic response of the disk by
%  % varying $\beta$.
%  We find that 
%  increasing $\beta$ even slightly away from    
%  zero dramatically reduce VSI growth rates. 
%  For vertically isothermal disks, we quantify the
%  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
%  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  is the radial temperature gradient, and 
%  $\gamma$ is the adiabatic index. Thus, in a thin disk the thermal timescale must be shorter than 
%  the dyamical timescale by $O(\epsilon)\ll 1$ for the VSI to operate.  
%  We apply our results  
%  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
%  estimate for the thermal timescale $\beta$ in the MMSN based on dust
%  opacity. In the inner parts of the MMSN, e.g. at $\lesssim 5$AU, 
%  we find the VSI would only operate on extremely short radial lengthscales, $\ll 
%  O(10^{-2}H)$, where $H$ is the disk scale height. On longer radial 
%  lengthscales of $O(10^{-2}H)$ to $O(10^{-1}H)$, we find the MMSN is
%  unstable to the VSI from $\sim 5$AU to $\sim 50$AU with a
%  characteristic growth time of $\sim 30$ local orbital periods.  
%  % ,and the dominant VSI mode in the MMSN is the `fundamental
%  % corrugation mode' which causes the disk column to oscillate
%  % vertically.
%  We thus suggest the VSI is at least dynamically important in the 
%  outer parts of protoplanetary disks at tens of AU. 
\end{abstract}

\section{Introduction}\label{intro}
Understanding the mechanisms of angular momentum transport is the
central problem in the theory of accretion disks 
\citep{lyndenbell74}. The   
magneto-rotational instability in the context of differentially
rotating disks \citep[MRI,][]{balbus91} is now accepted as a robust
pathway to magneto-hydrodynamic (MHD) turbulence, which transports
angular momentum outwards and thus enables mass accretion.

However, MHD turbulence may not be applicable to all  
accretion disks. For example, there may be regions in protoplanetary  
disks (PPDs) which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent MHD simulations
suggest significant portions of PPDs remain laminar \citep[e.g.][and references
therein]{lesur14,bai15,gressel15}. Thus, alternative mechanisms 
to angular momentum transport in PPDs, such as purely hydrodynamic
turbulence,  remain of interest \citep{armitage10}. In addition to
mass accretion, turbulence in PPDs will also
impact dust evolution and planetesimal formation \citep{chiang10}.

One possible route to hydrodynamic turbulence is the vertical shear
instability (VSI). The VSI is an 
axisymmetric linear instability 
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height $z$ from the disk midplane. 
Vertical shear occurs whenever the disk is baroclinic, i.e.\ when constant 
density and constant pressure surfaces are misaligned.  Baroclinicity, and thus vertical shear, is 
practically unavoidable in disks, except at special locations like the midplane.

The VSI  in disks is fundamentally equivalent to the
Goldreich-Schubert-Fricke instability 
\citep[GSFI,][]{goldreich67,fricke68} in differentially rotating stars.   
A height-dependence in $\Omega$, or vertical shear, can render
inertial waves unstable. \cite{barker15} gives a simple 
interpretation of the VSI: in an axisymmetric, inviscid, neutrally
stratified disk, the exchange of fluid parcels vertically, while
conserving their specific angular momentum, can result in a new
configuration of lower energy if the distribution of specific angular
momentum varies with $z$. The VSI is characterized by perturbations
with large vertical length-scales  relative to radial length-scales.

Early analysis of the VSI in radially and vertically local 
models of accretion disks \citep{urpin98,urpin03} showed that the VSI
requires a short thermal timescale so that perturbations evolve 
isothermally, unless the disk is neutrally stratified in the vertical
direction. This is because
vertical motions associated with the VSI 
are stabilized by buoyancy forces if the disk is sub-adiabatically
stratified (convectively stable). A sufficiently short thermal
timescale renders buoyancy ineffective, but a quantitative criteria is
desirable. This is our primary goal.  
%efficient exchang of heat with surroundings

The need for a short thermal timescale for the VSI to operate in
stably stratified disks has been demonstrated in recent numerical simulations 
\citep{nelson13,stoll14}. When the  thermal timescale $t_c$ is
parametrized in units of the orbital period  $P_\mathrm{orb}$,
\cite{nelson13} find that $t_c\lesssim  0.01P_\mathrm{orb}$ was needed
to observe the VSI. That is, the  disk temperature needs to remain
effectively unchanged. Radiation hydrodynamic simulations performed by
\cite{stoll14} also showed that the VSI is only sustained with the
inclusion of heating from stellar irradiation, which results in an
effective thermal timescale $t_c\ll P_\mathrm{orb}$ that maintains a
fixed temperature profile.

These studies show that the VSI is generally sensitive to disk
thermodynamics. It is thus important to quantify this dependence, for
example on disk parameters, in order to evaluate the relevance of the
VSI to realistic disks, which one may expect to be stably
stratified and have non-zero thermal timescales.  In order to determine
where in the disk the VSI may operate, and hence lead to hydrodynamic
turbulence, understanding the thermodynamic dependence of the linear VSI is essential. 
% since the transition
% Should a disk
% become turbulent through the VSI, it must first pass through the linear
% instability.

In this work we study the linear VSI in radially local, vertically
global models of astrophysical disks. This approach has
been adopted in recent studies of the linear VSI
\citep{nelson13,mcnally14,barker15}. However, these studies consider 
isothermal perturbations, for which the thermal timescale is formally
zero, and/or neutrally stratified disks in which buoyancy forces are
absent by construction. We account for both effects to study the
VSI in more general disk models. To do so, we examine 
sub-adiabatically stratified disks and include an energy equation 
with a source term that allows the thermodynamic response of 
perturbations to be varied smoothly between isothermal and 
adiabatic.

% Our models are also an extension of earlier 
% analysis by \cite{urpin98} and \cite{urpin03}, which accounts for thermal
% evolution, but considered vertically localized disturbances.
% Our 
% vertically global approach is necessary to study disturbances with
% vertical wavelengths comparable to or larger than the disk scale
% height. Such disturbances were found by \cite{nelson13} and
% \cite{stoll14} to dominate their numerical simulations. We thus focus
% on vertically global perturbations which, in fact, we find to be the
% dominant mode in protoplanetary disks with a realistic estimate for
% the thermodynamic timescales.

This paper is organized as follows. We first highlight in
\S\ref{vsi_require} the need of a very short thermal timescale for the
VSI to operate in stably stratified disks. There we also preview our
main result --- a quantitative thermodynamic criteria for the VSI ---
and a first application to protoplanetary disks. The remaining sections
are devoted to deriving these results from linear theory.

@@

>>>>

\title{Cooling Requirements for the Vertical Shear Instability 
  in Protoplanetary Disks}
%\title{Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula} 
\author{Min-Kai Lin \& Andrew Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

\begin{abstract}
  It is difficult to understand how cold circumstellar disks accrete onto their central stars.  Due to 
  lower levels of ionization, the magnetic transport mechanisms that operate in warmer disks are less effective. 
  A hydrodynamic mechanism, the vertical shear instability (VSI), offers a means to drive angular momentum 
  transport in cold accretion disks such as protoplanetary disks.  
  The VSI is driven by a weak vertical gradient in the disk's orbital motion.
  In order to grow, the VSI must overcome vertical buoyancy, a strong stabilizing influence in cold disks, where 
  heating is dominated by external irradiation.  Rapid cooling, via radiative losses, reduces the effective buoyancy
  and allows the VSI to operate.  In this paper, we quantify the cooling timescale, $t_c$, needed for growth of the VSI
  and consider the implications for protoplanetary disks where $t_c$ is determined by the opacity of dust 
  grains.  We perform a linear analysis of the VSI with cooling in vertically global and radially local disk models.
  For irradiated disks, we find that the VSI is most vigorous for rapid cooling with $t_c < \Omega_K^{-1} \epsilon |q| / (\gamma -1)$
  in terms of the Keplerian orbital frequency, $\OmK$; the disk's aspect ratio, $\epsilon \ll 1$; the radial 
  powerlaw temperature gradient, $q$; and the adiabatic index, $\gamma$.  For longer cooling times, the VSI 
  is much less effective because growth slows and shifts to smaller length scales, which are more prone to 
  viscous or turbulent decay.  We apply our results to a  Minimum Mass Solar Nebula (MMSN) disk model.  We
  find that the VSI is most effective at intermediate radii, from $\sim 5$AU to $\sim 50$AU with a
  characteristic growth time of $\sim 30$ local orbital periods.  Growth is suppressed by long cooling times both in the opaque 
  inner disk and the optically thin outer disk.  A reduction in the dust opacity by a factor of 10 increases cooling times 
  enough to quench the VSI at all disk radii.  Thus the formation of solid protoplanets, a sink for dust grains, can impede the VSI.
\end{abstract}

\section{Introduction}\label{intro}
Understanding how disks transport mass and angular momentum underlies 
many problems in astrophysics, including star and planet formation  
\citep[][]{lyndenbell74, armitage10}.  The turbulence associated with many transport mechanisms
is particularly important for dust evolution and planetesimal formation \citep{yl07, chiang10}.

Magneto-hydrodynamic (MHD) turbulence driven by
the magneto-rotational instability \citep[MRI,][]{balbus91} has long been the most
promising transport mechanism in low mass disks with weak self-gravity. 
However, many parts of protoplanetary  
disks (PPDs) are cold, have low levels of ionization, and do not support the MRI 
\citep{blaes94,salmeron03}. Recent simulations
suggest that significant portions of PPDs fail to develop MHD turbulence \citep[e.g.][]{simon13, lesur14,bai15,gressel15}.

A purely hydrodynamic mechanism could circumvent difficulties with non-ideal MHD, but must overcome
the strong centrifugal stability imposed the positive radial angular momentum gradient in nearly Keplerian disks \citep{balbus96}.
One possible route to hydrodynamic turbulence is the vertical shear
instability (VSI, \citealp{urpin98, nelson13}, hereafter N13).  The basic mechanism of the VSI  in disks was first
identified in the context of differentially rotating stars \citep[][]{goldreich67,fricke68}.   
The VSI arises when vertical shear, i.e.\ a variation in orbital motion along the rotation axis, 
destabilizes inertial-gravity waves, which are oscillations 
with rotation and buoyancy as restoring forces.  Vertical shear
occurs wherever the disk is baroclinic, i.e.\ when constant 
density and constant pressure surfaces are misaligned.  Baroclinicity, and thus vertical shear, is 
practically unavoidable in disks, except at special locations like the midplane.

To overcome centrifugal stabilization, the VSI triggers motions which are vertically elongated 
and radially narrow.  %This geometry minimizes the effects of the radial gradient of angular momentum.
Vertical elongation taps the free energy of the vertical shear \citep{umurhan13}, but is also subject to 
the stabilizing effects of vertical buoyancy. In principle, vertical buoyancy can be eliminated by strong internal heating, i.e.\ by the onset of convection.  
Real PPDs should be vertically stably stratified in the outer regions, beyond $\sim 1$--5 AU, where heating is 
irradiation dominated \citep{bitsch15}.  In the inner disk, strong vertical buoyancy is possible if accretion heating is weak
or is concentrated in surface layers \citep{gammie96,lesniak11}.

To overcome vertical buoyancy, the VSI requires a short cooling time (\citealp{goldreich67}; N13).  Rapid radiative cooling 
brings a moving fluid into thermal equilibrium with its surroundings, thereby diminishing buoyancy.  An isothermal equation
of state implies instantaneous cooling, and is the ideal context for studying the VSI (\citealp{urpin03}, N13, \citealp{barker15}).

A full understanding of the VSI in real disks requires considering finite cooling times, $t_{\rm c}$.
Simulations with a fixed, finite  $t_{\rm c}$ find that  VSI turbulence in stably stratified disks requires rapid cooling 
with $t_{\rm c} \lesssim 0.01/\Omega$, in terms of the orbital frequency $\Omega$ (N13).  N13 further find that the VSI can drive moderately strong transport, with Reynolds stresses of $\alpha \sim 10^{-3}$ times the mean pressure.    Simulations with realistic radiative transfer, in lieu of a constant $t_{\rm c}$ approximation,  find that the VSI in irradiated disks drive transport with $\alpha \sim 10^{-4}$
 in a  $\sim 2$ -- 10 AU PPD model \citep{stoll14}.

This paper presents the first comprehensive study of how finite cooling times affect the linear development of the VSI.  We consider models with constant $t_{\rm c}$  and  also with more realistic cooling from a dust opacity in both the optically thick (diffusive) and thin regimes, where appropriate.  Coupled with the results from numerical simulations, our results can be used to understand how the VSI operates over a wide range of model PPDs.  This paper is organized as follows. 
In \S\ref{vsi_require} we introduce and qualitatively explain our main result, the critical cooling time for the VSI.

\emph{TODO: the following is still a bit too wordy IMO but we can redo as a final step after these sections are finalized.}

@@

%The VSI is an axisymmetric linear instability 
%in three-dimensional (3D) disks in which the rotation frequency
%$\Omega$ depends on the height $z$ from the disk midplane. 
%Vertical shear occurs whenever the disk is baroclinic, i.e.\ when constant 
%density and constant pressure surfaces are misaligned.  Baroclinicity, and thus vertical shear, is 
%practically unavoidable in disks, except at special locations like the midplane.
%A height-dependence in $\Omega$, or vertical shear, can render
%inertial waves unstable. \cite{barker15} gives a simple 
%interpretation of the VSI: in an axisymmetric, inviscid, neutrally
%stratified disk, the exchange of fluid parcels vertically, while
%conserving their specific angular momentum, can result in a new
%configuration of lower energy if the distribution of specific angular
%momentum varies with $z$. The VSI is characterized by perturbations
%with large vertical length-scales  relative to radial length-scales.   
%Early analysis of the VSI in radially and vertically local 
%models of accretion disks \citep{urpin98,urpin03} showed that the VSI
%requires a short thermal timescale so that perturbations evolve 
%isothermally, unless the disk is neutrally stratified in the vertical
%direction. This is because
%vertical motions associated with the VSI 
%are stabilized by buoyancy forces if the disk is sub-adiabatically
%stratified (convectively stable). A sufficiently short thermal
%timescale renders buoyancy ineffective, but a quantitative criteria is
%desirable. This is our primary goal.  
%The need for a short thermal timescale for the VSI to operate in
%stably stratified disks has been demonstrated in recent numerical simulations 
%\citep{nelson13,stoll14}. When the  thermal timescale $t_c$ is
%parametrized in units of the orbital period  $P_\mathrm{orb}$,
%\cite{nelson13} find that $t_c\lesssim  0.01P_\mathrm{orb}$ was needed
%to observe the VSI. That is, the  disk temperature needs to remain
%effectively unchanged. Radiation hydrodynamic simulations performed by
%\cite{stoll14} also showed that the VSI is only sustained with the
%inclusion of heating from stellar irradiation, which results in an
%effective thermal timescale $t_c\ll P_\mathrm{orb}$ that maintains a
%fixed temperature profile.     
%These studies show that the VSI is generally sensitive to disk
%thermodynamics. It is thus important to quantify this dependence, for
%example on disk parameters, in order to evaluate the relevance of the
%VSI to realistic disks, which one may expect to be stably
%stratified and have non-zero thermal timescales.  In order to determine
%where in the disk the VSI may operate, and hence lead to hydrodynamic
%turbulence, understanding the thermodynamic dependence of the linear VSI is essential. 
%In this work we study the linear VSI in radially local, vertically
%global models of astrophysical disks. This approach has
%been adopted in recent studies of the linear VSI
%\citep{nelson13,mcnally14,barker15}. However, these studies consider 
%isothermal perturbations, for which the thermal timescale is formally
%zero, and/or neutrally stratified disks in which buoyancy forces are
%absent by construction. We account for both effects to study the
%VSI in more general disk models. To do so, we examine 
%sub-adiabatically stratified disks and include an energy equation 
%with a source term that allows the thermodynamic response of 
%perturbations to be varied smoothly between isothermal and 
%adiabatic.   
%This paper is organized as follows. We first highlight in
%\S\ref{vsi_require} the need of a very short thermal timescale for the
%VSI to operate in stably stratified disks. There we also preview our
%main result --- a quantitative thermodynamic criteria for the VSI ---
%and a first application to protoplanetary disks. The remaining sections
%are devoted to deriving these results from linear theory. 
%In \S\ref{setup} we list the governing equations and global
%equilibrium disk models under   
%consideration. We formulate the linear stability problem in 
%\S\ref{linear} and simplify it by making the radially local
%approximation. We study the linear problem using two approaches. In
%\S\ref{analytical} we present an analytical discussion of vertically
%isothermal disks. Our treatment is grossly simplified owing to a
%number of approximations made, but closely parallels previous work, 
%allowing us to highlight the essential effects of vertical shear, and to
%derive a characteristic upper limit to the thermal timescale for the
%VSI. This timescale is \emph{much shorter} than the dynamical
%timescale. In \S\ref{numerical} we numerically
%solve the linear problem in the radially local approximation, but
%relax all other approximations made in our analytical approach.    
%We find basic agreement with our analysis, including
%the aforementioned characteristic thermal timescale. 
%In
%\S\ref{application} we apply our results to the Minimum Mass
%Solar Nebula (MMSN) by modeling thermal 
%timescales based on realistic disk properties, and find that the VSI
%can operate effectively in the outer parts of the MMSN. We summarize in
%\S\ref{summary} with a discussion of caveats and possible directions
%for future work.
